Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-06-21 paper.tex
More fixes!

\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{listings}
\usepackage{subcaption}
\usepackage{makecell}

@@

\subsection{The Scan Primitive}
\label{sec:the-scan-primitive}
The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{5219801,10.5555/1098666}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). To allow parallelization, \( \oplus \) must be associative, but it is not necessarily commutative, as demonstrated in structures like the \emph{bicyclic semigroup}~\cite{Clifford1961}\footnote{Despite its name, this is a monoid.} and Rabin-Karp's \emph{second fingerprinting method for string matching}~\cite[Section 6]{Karp:1987:ERP}. In a scan, the result at the \( n \)-th element is the reduction of the preceding subsequence of elements. If the subsequence includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1]
\]
\[
  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]
\[
  \text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
\]
Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science. Harris~\cite{1292373} offers a taxonomy that relates depth, fanout, and wire tracks, while Hinze~\cite{10.1007/978-3-540-27764-4_11} develops an algebraic framework for scans. Snir~\cite{10.1016/0196-67748690003-9} proved that depth $d$ and size $s$ are related by $s + d \ge 2n - 2$, and Fich~\cite{10.1145/800061.808738} proved that among minimum depth scans Ladner-Fischer~\cite{10.1145/322217.322232} networks have optimal size. Blelloch~\cite{Blelloch:1989:SAP} adapted scan to the PRAM model and popularized its use as an algorithmic primitive. Finally, Merrill and Garland~\cite{Merrill2016} provide a review of GPU scan implementations, complementing earlier work by Merrill and Grimshaw~\cite{Merrill2009}.

@@

  \makebox[\textwidth][c]{
    \begin{subfigure}[c]{0.48\textwidth}
      \centering
      \includegraphics[height=6cm,keepaspectratio]{graphics/StreamScan.pdf}
      \caption{\emph{Stream Scan.}}
      \label{fig:stream-scan}
    \end{subfigure}
    \hfill
    \begin{subfigure}[c]{0.48\textwidth}
      \centering
      \includegraphics[height=6cm,keepaspectratio]{graphics/DecoupledLookback.pdf}
      \caption{\emph{Decoupled Lookback.}}
      \label{fig:decoupled-lookback}
    \end{subfigure}%
  }

@@

\subsection{Ideal Scan Architecture}
In terms of performance, an ideal global scan algorithm should minimize global memory movement, fully utilize the device’s memory bandwidth, and avoid costly global barriers. By these criteria alone, the \emph{Chained Scan} architecture emerges as the preferred solution; moreover, its single-kernel design offers compelling portability advantages across diverse hardware. By eliminating global barriers, \emph{Chained Scan} naturally conforms to the GPU’s oversubscription model, allowing occupancy control to be delegated the hardware's workgroup scheduler. In contrast, hierarchical approaches such as \emph{Reduce-then-Scan} require workgroup raking and fixed-occupancy persistent threads to bound the spine’s size, necessitating explicit kernel launches and manual tuning of launch configurations to maintain optimal occupancy.

Hardware-level considerations underscore the limitations of the fixed-launch approach. On GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—fixed launches can lead to performance degradation due to cache thrashing. This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups. Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns—or applications like radix sorting, where the scattering phase of each pass results in $\textit{radixDigits} \cdot o$ unique memory locations—may experience significant performance degradation.

Moreover, the fused nature of \emph{Chained Scan} enables in-situ compaction: algorithms requiring compaction can merge that work directly into the scan kernel, avoiding redundant writes or secondary compaction passes. This fusion benefits applications ranging from data deduplication to compression. However, \emph{Chained Scan} has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple GPUs do not. Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite starvation, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
\emph{Chained Scans} are \emph{one-way producer-consumer} algorithms; each workgroup consumes reductions only from lower-ID counterparts, thereby enforcing the ordering required to elide barriers. For such algorithms, guaranteeing execution of the lowest-ID workgroup is sufficient to ensure termination. If subgroups from the lowest-ID workgroup are not guaranteed execution, dependent workgroups may spin indefinitely, waiting for updates that never arrive. A \emph{forward-progress guarantee} ensures that all occupant subgroups---potentially from different workgroups---are eventually granted execution, thus preventing indefinite starvation.

The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\label{sec:tdr}\footnote{Operating systems handle unresponsive programs differently. On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds. Other operating systems—such as macOS and Linux—employ their own timeout mechanisms with varying behaviors. For consistency, we adopt Windows’s 2-second limit as a reasonable expectation for maximum execution time.} (TDR), crashing the program. Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.

@@

\begin{figure*}[t]
  \centering
  \includegraphics[width=.95\textwidth]{graphics/DecoupledFallback.pdf}
  \caption{\emph{Decoupled Fallback}. For brevity, the chained propagation phase is condensed; see Figure~\ref{fig:decoupled-lookback} for the full behavior.}
  \label{fig:decoupled-fallback}
\end{figure*}

@@

\section{Decoupled Fallback}
\emph{Decoupled Fallback} addresses both inefficiencies. Our key advance is to utilize the parallelism available in the workgroup to perform the fallback computation, rather than performing it sequentially. To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained Propagation} logic is typically performed by the first thread.

@@

\section{Implementation}
At a high level, our algorithm follows a five-phase construction:
\begin{enumerate}
  \item[(0)] \textbf {Initialization}: The first thread in the workgroup atomically acquires its work-tile by atomically incrementing a value in global memory.
  \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering. For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$.
  \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking Kogge-Stone~\cite{5009159} subgroup scans,
        allowing the subgroup-block scan to remain barrier-free and in registers. The subgroup-block reduction is then posted to shared memory. For a workgroup of size $w$, subgroup-raking reduces the size of the local spine to $w/s$.
  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion.
  \item \textbf{Chained Propagation, Decoupled Fallback}: The local reduction is posted to global memory. Then, the first subgroup in the workgroup performs a lookback operation to acquire the reduction result of all previous workgroups, engaging in fallback operations as necessary. Once completed, this subgroup posts the results to shared memory for the workgroup to consume.
  \item \textbf{Write Phase}: Each subgroup incorporates the results from the workgroup-wide scan and lookback operation to produce and write out the correct output.
\end{enumerate}
Phase 0 ensures that the work tiles are consumed in serial order. Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy. Phase 4, \emph{Decoupled Fallback}, is responsible for managing all inter-workgroup dependencies. As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations.

@@

\subsection{Workgroup-Wide Scan}
The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches. To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{10.1145/322217.322232} scan network embedded with Kogge-Stone~\cite{5009159} subgroup scans to perform the scan across the local spine. Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts. To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$. Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction. Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index. However, such access is interpreted by the hardware as a subgroup \emph{broadcast} operation, resulting in no conflicts. Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking.

@@

      \If{$\textit{thread\_id} < \textit{step}$}{
        $\textit{temp} \gets \text{subgroupKoggeStoneScan}(\textit{x}[\textit{thread\_id} + \textit{top\_offset}])$\;
        $\textit{x}[\textit{thread\_id} + \textit{top\_offset}] \gets \textit{temp}$\;
        \If{$\text{\normalfont highestRankLane}(\textit{thread\_id})$}{
          $\textit{x}[(\textit{thread\_id} / \textit{S}) + \textit{step} + \textit{top\_offset}] \gets \textit{temp}$\;
        }
      }

@@

        $\textit{cond1} \gets \textit{fanout\_index} < \textit{spine\_length}$\;
        $\textit{cond2} \gets (\textit{fanout\_index} \,\&\, (\textit{j} - 1)) \geq \textit{reduced\_stride}$\;

@@

Fallbacks require a tile state representation and an atomic operation that can safely transition a tile to \emph{Ready} without overwriting a more advanced state. When an update attempt is made, a blocking tile may be in any of the three states due to either of two scenarios: (1) a false-positive fallback on a tile that transitions to \emph{Ready} or \emph{Inclusive} before fallback completion, or (2) another workgroup completing its fallback first, leaving the tile in \emph{Ready}. If the tile state is \emph{Inclusive} then the write must be skipped, especially because races in writes for the parts may leave the tile permanently in an unreadable state. Thus, a simple atomic store is not suitable. An atomic compare-and-swap would have the correct semantics, but WebGPU lacks this primitive (it does have \emph{atomicCompareExchangeWeak}, which could possibly be made to work but makes reasoning about correctness and performance more difficult). Instead, we encode the tile state in the most significant bits of the atomic value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}. This allows an \emph{atomicMaximum}, which successfully updates to a more advanced state but prevents a state regression, to correctly perform updates.

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. We find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum. Our test hardware, shown in Figure~\ref{tab:hardware}, spans a wide range of FPG and performance capabilities. We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing.
\begin{table*}
  \centering
  \begin{tabular}{l l c r r}
    \toprule
    Vendor & Device Name            & FPG         & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620        & \textbf{Yes} & 9         & 8--32         \\
    ARM    & Mali-G78 MP20          & \textbf{No}  & 40        & 16            \\
    Apple  & M3     (10 Cores)      & \textbf{No}  & 80        & 32            \\
    Apple  & M1 Max (32 Cores)      & \textbf{No}  & 160       & 32            \\
    NVIDIA & GeForce RTX 2080 Super & \textbf{Yes} & 192       & 32            \\
    AMD    & Radeon RX 7900 XT      & \textbf{Yes} & 420       & 32--64        \\
    \bottomrule
  \end{tabular}
  \caption{Hardware Specifications of Testing Devices; FPG determinations are based on the data collected by Sorensen et al.~\cite{sorensen2021}.\label{tab:hardware}}
\end{table*}

@@

Cross-API comparisons remain challenging, as our CUDA implementation consistently outperforms its Dawn counterpart despite near-identical high-level implementations. The performance cost of \emph{DF} is most pronounced in CUDA when the entire input fits within the 2080 Super's 4~MB L2 cache, where reduced global memory latency exposes \emph{DF's} increased synchronization and compute overhead compared to CUB\@. However, at larger input sizes, where memory accesses exceed cache capacity, \emph{DF} reaches parity with CUB and achieves \emph{Memcpy} speeds.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{graphics/cuda_plot.pdf}
  \caption{Performance versus CUB (NVIDIA 2080 Super).}
\end{figure}

@@

\subsubsection{Same-API Performance Versus Baseline Measures}
We next compare \emph{DF} against \emph{Reduce-then-Scan} (\emph{RTS}), which serves as a baseline to highlight the performance improvements of our approach over the prior FPG-constrained technique. This evaluation is conducted using Dawn across the full range of tested devices.

@@

\subsection{Decoupled Fallback Statistics}
To better understand how devices interact with \emph{DF}, we collect the following statistics during its execution:

\label{sec:Sim-Blocking}
To demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally omitting work tile reduction updates. Because these updates are skipped, the immediate successor workgroup reads \emph{Not Ready}, spinning and triggering a fallback. Starting with infrequent omissions---1 in 512 work tiles---we progressively increase the omission rate to near-continuous levels---1 in 2 work tiles---recording \emph{DF} statistics at each step.
\begin{figure*}
  \centering
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/fallbacksInitiated_plot.pdf}
    \caption{Total number of fallback operations triggered as simulated blocking rates increase. This statistic verifies that the simulation effectively induces fallbacks as intended.\label{fig:fallbacks_initiated}}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/totalSpins_plot.pdf}
    \caption{Average number of waiting spins per workgroup as simulated blocking rates increase. Higher spin counts correlate with increased fallback operations, reinforcing the behavior observed in Figure~\ref{fig:fallbacks_initiated}.\label{fig:total_spins}}
  \end{subfigure}

@@

@@

Furthermore, this data reveals two notable behaviors: (1) As demonstrated by the Apple M3, a scheduler can be fair yet still lack FPG, demonstrating that fairness and FPG are not necessarily correlated. (2) The ARM Mali-G78 MP20 exhibits the most unfair scheduling among all tested devices, with a fallback rate of 57.8\% relative to workgroups launched. However, the average number of spins per workgroup is only 2.52---below our fallback trigger threshold of 4 spins. This suggests a bimodal distribution of waiting times, where workgroups either proceed immediately with minimal delay or experience prolonged stalls that trigger fallbacks.

>>>>

\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{listings}
\usepackage{subcaption}
\usepackage{makecell}
\usepackage{stfloats}

@@

\subsection{The Scan Primitive}
\label{sec:the-scan-primitive}
The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{5219801,10.5555/1098666}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). To allow parallelization, \( \oplus \) must be associative, but it is not necessarily commutative, as demonstrated in structures like the \emph{bicyclic semigroup}~\cite{Clifford1961}\footnote{Despite its name, this is a monoid.} and Rabin-Karp's \emph{second fingerprinting method for string matching}~\cite[Section 6]{Karp:1987:ERP}. In a scan, the result at the \( n \)-th element is the reduction of the preceding subsequence of elements. If the subsequence includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1]
\]
\[
  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]
\[
  \text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
\]
Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science. Harris~\cite{1292373} offers a taxonomy that relates depth, fanout, and wire tracks, while Hinze~\cite{10.1007/978-3-540-27764-4_11} develops an algebraic framework for scans. Snir~\cite{10.1016/0196-67748690003-9} proved that depth $d$ and size $s$ are related by $s + d \ge 2n - 2$, and Fich~\cite{10.1145/800061.808738} proved that among minimum depth scans Ladner-Fischer~\cite{10.1145/322217.322232} networks have optimal size. Blelloch~\cite{Blelloch:1989:SAP} adapted scan to the PRAM model and popularized its use as an algorithmic primitive. Merrill and Garland~\cite{Merrill2016} provide a review of GPU scan implementations, complementing earlier work by Merrill and Grimshaw~\cite{Merrill2009}.

@@

  \makebox[\textwidth][c]{
    \begin{subfigure}[c]{0.48\textwidth}
      \centering
      \includegraphics[height=6cm,keepaspectratio]{graphics/StreamScan.pdf}
      \caption{\emph{StreamScan.}}
      \label{fig:stream-scan}
    \end{subfigure}
    \hfill
    \begin{subfigure}[c]{0.48\textwidth}
      \centering
      \includegraphics[height=6cm,keepaspectratio]{graphics/DecoupledLookback.pdf}
      \caption{\emph{Decoupled Lookback.}}
      \label{fig:decoupled-lookback}
    \end{subfigure}%
  }

@@

\subsection{Ideal Scan Architecture}
In terms of performance, an ideal global scan algorithm should minimize global memory movement, fully utilize the device’s memory bandwidth, and avoid costly global barriers. By these criteria alone, the \emph{Chained Scan} architecture emerges as the preferred solution; moreover, its single-kernel design offers compelling portability advantages across diverse hardware. By eliminating global barriers, \emph{Chained Scan} naturally conforms to the GPU’s oversubscription model, allowing occupancy control to be delegated to the hardware's workgroup scheduler. In contrast, hierarchical approaches such as \emph{Reduce-then-Scan} require workgroup raking and fixed-occupancy persistent threads to bound the spine’s size, necessitating manual tuning of launch configurations to maintain optimal occupancy.

Hardware-level considerations underscore the limitations of the fixed-launch approach. On GPUs where the global memory data cache is indexed by physical addresses---and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)---fixed launches can lead to performance degradation due to cache thrashing. This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups. Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns---or applications like radix sorting, where the scattering phase of each pass results in $\textit{radixDigits} \cdot o$ unique memory locations---may experience significant performance degradation.

Moreover, the fused nature of \emph{Chained Scan} enables \emph{in-situ compaction}: algorithms requiring compaction can merge that work directly into the scan kernel, avoiding redundant writes or secondary compaction passes. This fusion benefits applications ranging from data deduplication to compression. However, \emph{Chained Scan} has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple GPUs do not. Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite starvation, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
\emph{Chained Scans} are \emph{one-way producer-consumer} algorithms; each workgroup consumes reductions only from \emph{lower-rank}\footnote{We specifically use the term \emph{rank} here to generalize between which ID---atomically acquired or scheduler-assigned---is used to determine this ordering. This distinction matters because the choice of ID determines which progress model, as defined by Sorensen et al.~\cite{sorensen2021}, is required to guarantee termination. Working under the assumption that the scheduler-assigned ID is used, Sorensen et al.~\cite{sorensen2021, sorensen2018} classify \emph{Chained Scan} as requiring \emph{HSA} (Heterogeneous System Architecture), a model that guarantees progress for the lowest-ranked workgroup. However, if the atomically acquired ID determines rank, then the HSA guarantee no longer suffices, as the lowest scheduler-assigned ID may not correspond to the lowest atomic ID. In this case, termination requires only \emph{OBE} (Occupancy-Bound Execution), which guarantees progress once a workgroup begins execution.} counterparts, thereby enforcing the ordering required to elide barriers. For such algorithms, guaranteeing execution of the lowest-rank workgroup is sufficient to ensure termination. If subgroups from the lowest-rank workgroup are not guaranteed execution, dependent workgroups may spin indefinitely, waiting for updates that never arrive. A \emph{forward-progress guarantee}, in the sense required by \emph{Chained Scan}, ensures that all occupant subgroups---potentially from different workgroups---are eventually granted execution after becoming active, thus preventing indefinite starvation.

The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Detection and Recovery}\label{sec:tdr}\footnote{Operating systems handle unresponsive programs differently. On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds. Other operating systems---such as macOS and Linux---employ their own timeout mechanisms with varying behaviors. For consistency, we adopt Windows’s 2-second limit as a reasonable expectation for maximum execution time.} (TDR), crashing the program. Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on their target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing the advantages of \emph{Chained Scan} for broader compatibility.

@@

@@

\begin{figure*}
  \centering
  \includegraphics[width=.95\textwidth]{graphics/DecoupledFallback.pdf}
  \caption{\emph{Decoupled Fallback}. For brevity, the lookback is condensed; see Figure~\ref{fig:decoupled-lookback} for the full behavior, which is shared with \emph{Decoupled Lookback}.}
  \label{fig:decoupled-fallback}
\end{figure*}
\section{Decoupled Fallback}
\emph{Decoupled Fallback} addresses both inefficiencies. Our key advance is to utilize the parallelism available in the workgroup to perform the fallback computation, rather than performing it sequentially. To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained Propagation} logic is typically performed by the first thread.

@@

\section{Implementation}
At a high level, our algorithm follows a five-phase construction:
\begin{enumerate}
  \item[(0)] \textbf {Initialization}: The first thread of the workgroup atomically acquires its work-tile by atomically incrementing a value in global memory.
  \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering. For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$.
  \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking Kogge-Stone~\cite{5009159} subgroup scans,
        allowing the subgroup-block scan to remain barrier-free and in registers. The subgroup-block reduction is then posted to shared memory. For a workgroup of size $w$, subgroup-raking reduces the size of the local spine to $w/s$.
  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion.
  \item \textbf{Chained Propagation, Decoupled Fallback}: The local reduction is posted to global memory. Then, the first subgroup in the workgroup performs a lookback operation to acquire the reduction result of all previous workgroups, engaging in fallback operations as necessary. Once completed, this subgroup posts the results to shared memory for the workgroup to consume.
  \item \textbf{Write Phase}: Each subgroup incorporates the results from the workgroup-wide scan and lookback operation to produce and write out the correct output.
\end{enumerate}
Phase 0 ensures that the work tiles are consumed in serial order. Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy. Phase 4, \emph{Decoupled Fallback}, is responsible for managing all inter-workgroup dependencies. As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations.

@@

\subsection{Workgroup-Wide Scan}
The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches. To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{10.1145/322217.322232} scan network embedded with Kogge-Stone~\cite{5009159} subgroup scans to perform the scan across the local spine. Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts. To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network---in our case, $s$. Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction. Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index. However, such access is interpreted by the hardware as a subgroup \emph{broadcast} operation, resulting in no conflicts. Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking.

@@

      \If{$\textit{thread\_id} < \textit{step}$}{
        $\textit{temp} \gets \text{subgroupKoggeStone}(\textit{x}[\textit{thread\_id} + \textit{top\_offset}])$\;
        $\textit{x}[\textit{thread\_id} + \textit{top\_offset}] \gets \textit{temp}$\;
        \If{$\text{\normalfont highestRankLane}(\textit{thread\_id})$}{
          $\textit{x}[(\textit{thread\_id} / \textit{S}) + \textit{step} + \textit{top\_offset}] \gets \textit{temp}$\;
        }
      }

@@

        $\textit{cond1} \gets \textit{fanout\_index} < \textit{spine\_length}$\;
        $\textit{cond2} \gets \textit{fanout\_index} \bmod \textit{j} \geq \textit{reduced\_stride}$\;

@@

Fallbacks require a tile state representation and an atomic operation that can safely transition a tile to \emph{Ready} without overwriting a more advanced state. When an update attempt is made, a blocking tile may be in any of the three states due to either of two scenarios: (1) a false-positive fallback on a tile that transitions to \emph{Ready} or \emph{Inclusive} before fallback completion, or (2) another workgroup completing its fallback first, leaving the tile in \emph{Ready}. If the tile state is \emph{Inclusive} then the write must be skipped, especially because races in writes for the parts may leave the tile permanently in an unreadable state. Thus, a simple atomic store is not suitable. An atomic compare-and-swap would have the correct semantics, but WebGPU lacks this primitive (it does have \emph{atomicCompareExchangeWeak}, which could possibly be made to work but makes reasoning about correctness and performance more difficult). Instead, we encode the tile state in the most significant bits of the atomic value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}. This allows an \emph{atomicMaximum} to successfully update to a more advanced state while preventing state regression.

@@

\section{Evaluation}
\subsection{Experimental Setup}
We configure workgroups with a size $w = 256$ threads, where each thread processes $p = 4$ vectors of size 4, resulting in 16 elements per thread and a total work tile size of $t = 4096$. We find a maximum spin count of 4 effectively balances responsiveness to blocking while minimizing false-positive fallbacks. As the size of the local spine is inversely proportional to the subgroup size, we allocate sufficient shared memory to accommodate the WebGPU standard's minimum subgroup size of 4. Due to our overall thriftiness with shared memory, total usage remains modest at only $\sim$1 KB\@. For all tests, we use an input size of $2^{25}$ and an inclusive prefix sum. Our test hardware, shown in Table~\ref{tab:hardware}, spans a wide range of FPG and performance capabilities. We empirically find that Google Chrome's native implementation of WebGPU, Dawn, provides the best performance and use it for all testing.
\begin{table*}
  \centering
  \begin{tabular}{l l c r r}
    \toprule
    Vendor & Device Name            & FPG         & Occupancy & Subgroup Size \\
    \midrule
    Intel  & HD Graphics 620          & \textbf{Yes} & 9         & 8--32         \\
    ARM    & Mali-G78 MP20 (Pixel 6a) & \textbf{No}  & 40        & 16            \\
    Apple  & M3     (10 Cores)        & \textbf{No}  & 80        & 32            \\
    Apple  & M1 Max (32 Cores)        & \textbf{No}  & 160       & 32            \\
    NVIDIA & GeForce RTX 2080 Super   & \textbf{Yes} & 192       & 32            \\
    AMD    & Radeon RX 7900 XT        & \textbf{Yes} & 420       & 32--64        \\
    \bottomrule
  \end{tabular}
  \caption{Hardware Specifications of Testing Devices; FPG determinations are based on the data collected by Sorensen et al.~\cite{sorensen2021}.\label{tab:hardware}}
\end{table*}

\begin{figure}[t]
  \centering
  \includegraphics[width=\linewidth]{graphics/cuda_plot.pdf}
  \caption{Performance versus CUB (NVIDIA 2080 Super).}
  \label{fig:cuda}
\end{figure}

@@

As can be seen in Figure \ref{fig:cuda}, cross-API comparisons remain challenging, as our CUDA implementation consistently outperforms its Dawn counterpart despite near-identical high-level implementations. The performance cost of \emph{DF} is most pronounced in CUDA when the entire input fits within the 2080 Super's 4~MB L2 cache, where reduced global memory latency exposes \emph{DF's} increased synchronization and compute overhead compared to CUB\@. However, at larger input sizes, where memory accesses exceed cache capacity, \emph{DF} reaches parity with CUB and achieves \emph{Memcpy} speeds.

@@

\subsubsection{Same-API Performance Versus Baseline Measures}
\label{sec:SameVsBase}
We next compare \emph{DF} against \emph{Reduce-then-Scan} (\emph{RTS}), which serves as a baseline to highlight the performance improvements of our approach over the prior FPG-constrained technique. This evaluation is conducted using Dawn across the full range of tested devices.

@@

\subsection{Decoupled Fallback Statistics}
\begin{figure*}
  \centering
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/fallbacksInitiated_plot.pdf}
    \caption{Total number of fallback operations triggered as simulated blocking rates increase. This statistic verifies that the simulation effectively induces fallbacks as intended.\label{fig:fallbacks_initiated}}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/totalSpins_plot.pdf}
    \caption{Average number of waiting spins per workgroup as simulated blocking rates increase. Higher spin counts correlate with increased fallback operations, reinforcing the behavior observed in Figure~\ref{fig:fallbacks_initiated}.\label{fig:total_spins}}
  \end{subfigure}

@@

To better understand how devices interact with \emph{DF}, we collect the following statistics during its execution:

@@

This data reveals two notable behaviors: (1) As demonstrated by the Apple M3, a scheduler can be fair yet still lack FPG, demonstrating that fairness and FPG are not necessarily correlated. (2) The ARM Mali-G78 MP20 exhibits the most unfair scheduling among all tested devices, with a fallback rate of 57.8\% relative to workgroups launched. However, the average number of spins per workgroup is only 2.52---below our fallback trigger threshold of 4 spins. This suggests a bimodal distribution of waiting times, where workgroups either proceed immediately with minimal delay or experience prolonged stalls that trigger fallbacks.

\subsection{Simulated Blocking}%
\label{sec:Sim-Blocking}
To demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally omitting work tile reduction updates. Because these updates are skipped, the immediate successor workgroup reads \emph{Not Ready}, spinning and triggering a fallback. Starting with infrequent omissions---1 in 512 work tiles---we progressively increase the omission rate to near-continuous levels---1 in 2 work tiles---recording \emph{DF} statistics at each step.
